---
title: "Salmon Project"
author: "Carissa Mayo"
date: "2021-11-18"
output: pdf_document
---

# SALMON POPULATION ANALYSIS - COMPLETE WORKFLOW
### Analysis of factors affecting Northwestern salmon populations

# Load required libraries

```{r}
library(tidyverse)
library(plyr)
library(car)  # For VIF calculations
library(here)
library(ggplot2)
```

## SECTION 1: DATA IMPORT AND INITIAL SETUP


Read primary datasets (adjust file paths as needed)
Data was obtained from NOAA and research was done to determine number of dams for each river.
```{r}
salmon_data <- read.csv("slcws.csv")
rainfall_data <- read.csv("par.csv")
logging_data <- read.csv("logging.csv")
consumption_data <- read.csv("usconsumption.csv")
streamflow_data <- read.csv("bonnevillestreamflow.csv")

# Read water quality data - each parameter has its own CSV with all rivers
wq_files <- c("dissolvedoxygen.csv", "pH.csv", "watertemperature.csv", 
              "chloride.csv", "lead.csv", "chromium.csv", 
              "ammonia.csv", "arsenic.csv")

wq_data <- list()
for (file in wq_files) {
  param_name <- gsub(".csv", "", file)
  if (file.exists(file)) {
    wq_data[[param_name]] <- read.csv(file)
  }
}
```

# SECTION 2: DATA CLEANING AND STANDARDIZATION

Clean salmon data
```{r}
# Filtering out rivers not in Columbia Basin
clean_salmon_data <- function(df) {
  df %>%
    dplyr::rename(Location = Common.Population.Name)
}

salmon_clean <- clean_salmon_data(salmon_data)
```

# SECTION 3: AGGREGATE WATER QUALITY DATA

Calculate yearly averages for water quality parameters
```{r}
# Aggregate water quality data
aggregate_wq <- function(wq_list) {
  aggregated <- list()
  
  clean_location <- function(df) {
    df %>%
      mutate(
        Location = case_when(
          Location == "Umatilla" ~ "Umatilla River",
          TRUE ~ Location
        )
      )
  }
  
  # Dissolved Oxygen
  if (!is.null(wq_list$dissolvedoxygen)) {
    aggregated$DO <- wq_list$dissolvedoxygen %>%
      clean_location() %>%
      dplyr::group_by(Location, Year) %>%
      dplyr::summarize(DissolvedOxygen = mean(Value, na.rm = TRUE), .groups = "drop")
  }
  
  # pH
  if (!is.null(wq_list$pH)) {
    aggregated$pH <- wq_list$pH %>%
      clean_location() %>%
      dplyr::group_by(Location, Year) %>%
      dplyr::summarize(pH = mean(Value, na.rm = TRUE), .groups = "drop")
  }
  
  # Water Temperature (multiple statistics)
  if (!is.null(wq_list$watertemperature)) {
    aggregated$temp_avg <- wq_list$watertemperature %>%
      clean_location() %>%
      dplyr::group_by(Location, Year) %>%
      dplyr::summarize(Temp = mean(Value, na.rm = TRUE), .groups = "drop")
    
    aggregated$temp_max <- wq_list$watertemperature %>%
      clean_location() %>%
      dplyr::group_by(Location, Year) %>%
      dplyr::summarize(TempMax = max(Value, na.rm = TRUE), .groups = "drop")
    
    aggregated$temp_min <- wq_list$watertemperature %>%
      clean_location() %>%
      dplyr::group_by(Location, Year) %>%
      dplyr::summarize(TempMin = min(Value, na.rm = TRUE), .groups = "drop")
    
    aggregated$temp_summer <- wq_list$watertemperature %>%
      clean_location() %>%
      filter(Month >= 6 & Month <= 8) %>%
      dplyr::group_by(Location, Year) %>%
      dplyr::summarize(TempSummer = mean(Value, na.rm = TRUE), .groups = "drop")
  }
  
  # Chloride
  if (!is.null(wq_list$chloride)) {
    aggregated$chloride <- wq_list$chloride %>%
      clean_location() %>%
      dplyr::group_by(Location, Year) %>%
      dplyr::summarize(Chloride = mean(Value, na.rm = TRUE), .groups = "drop")
  }
  
  # Ammonia
  if (!is.null(wq_list$ammonia)) {
    aggregated$ammonia <- wq_list$ammonia %>%
      clean_location() %>%
      dplyr::group_by(Location, Year) %>%
      dplyr::summarize(Ammonia = mean(Value, na.rm = TRUE), .groups = "drop")
  }
  
  # Lead
  if (!is.null(wq_list$lead)) {
    aggregated$lead <- wq_list$lead %>%
      clean_location() %>%
      dplyr::group_by(Location, Year) %>%
      dplyr::summarize(Lead = mean(Value, na.rm = TRUE), .groups = "drop")
  }
  
  # Arsenic
  if (!is.null(wq_list$arsenic)) {
    aggregated$arsenic <- wq_list$arsenic %>%
      clean_location() %>%
      dplyr::group_by(Location, Year) %>%
      dplyr::summarize(Arsenic = mean(Value, na.rm = TRUE), .groups = "drop")
  }
  
  # Chromium
  if (!is.null(wq_list$chromium)) {
    aggregated$chromium <- wq_list$chromium %>%
      clean_location() %>%
      dplyr::group_by(Location, Year) %>%
      dplyr::summarize(Chromium = mean(Value, na.rm = TRUE), .groups = "drop")
  }
  
  return(aggregated)
}

wq_aggregated <- aggregate_wq(wq_data)
```


# SECTION 4: PREPARE AUXILIARY DATASETS

```{r}
# Process streamflow data
streamflow_clean <- streamflow_data %>%
  dplyr::rename(Year = Water.Year) %>%
  mutate(
    Year = as.numeric(Year),
    MaxFlow = apply(dplyr::select(., Oct:Sep), 1, max, na.rm = TRUE),
    MinFlow = pmin(Oct, Nov, Dec, Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, na.rm = TRUE),
    SDWinterFlow = apply(dplyr::select(., Dec, Jan, Feb), 1, sd, na.rm = TRUE)
  ) %>%
  dplyr::select(Year, MaxFlow, MinFlow, SDWinterFlow)

# Process logging data
logging_clean <- logging_data %>%
  mutate(
    Idaho          = as.numeric(gsub("[^0-9.-]", "", Idaho)),
    Oregon.b       = as.numeric(gsub("[^0-9.-]", "", Oregon.b)),
    Washington.b   = as.numeric(gsub("[^0-9.-]", "", Washington.b)),
    TotalNWLogging = Idaho + Oregon.b + Washington.b
  ) %>%
  dplyr::select(Year, TotalNWLogging, OregonLogging = Oregon.b, 
         WashingtonLogging = Washington.b, IdahoLogging = Idaho)

# Process consumption data
consumption_clean <- consumption_data %>%
  dplyr::select(Year, FishConsumption = Consumption)

# Process rainfall data
rainfall_clean <- rainfall_data %>%
  dplyr::select(Year, Yearly.Rainfall)
```

# SECTION 5: MERGE ALL DATASETS

```{r}
# Start with salmon data and progressively merge
merged_data <- salmon_clean %>%
  left_join(streamflow_clean, by = "Year") %>%
  left_join(logging_clean, by = "Year") %>%
  left_join(consumption_clean, by = "Year") %>%
  left_join(rainfall_clean, by = "Year")

# Merge water quality data
for (wq_name in names(wq_aggregated)) {
  merged_data <- merged_data %>%
    left_join(wq_aggregated[[wq_name]], by = c("Location", "Year"))
}

# Remove duplicate rows
merged_data <- distinct(merged_data)
```

# SECTION 6: REGRESSION ANALYSIS
```{r}
overall_model <- lm(Spawners ~ totaldams + colsnakedams + MaxFlow + MinFlow + 
                      SDWinterFlow + TotalNWLogging + FishConsumption + 
                      Yearly.Rainfall + DissolvedOxygen + pH + Temp + 
                      TempSummer + Ammonia + Chloride,
                    data = merged_data)
summary(overall_model)

```
```{r}
ggplot(merged_data, aes(x = colsnakedams, y = Spawners)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "blue", linewidth = 1) +
  labs(
    title = "Relationship Between Spawners and Number of Columbia–Snake River Dams",
    x = "Number of Dams (colsnakedams)",
    y = "Spawner Count"
  ) +
  theme_minimal(base_size = 14)
```
### Each additional dam is associated with ~428 fewer returning spawners, and the effect is statistically significant (p < 0.01).

# SECTION 7: INDIVIDUAL RIVER ANALYSES
```{r}
analyze_river <- function(data, river_name, predictors) {
  cat("\n", rep("=", 80), "\n", sep = "")
  cat("ANALYSIS FOR:", river_name, "\n")
  cat(rep("=", 80), "\n\n", sep = "")
  
  river_data <- data %>% filter(Location == river_name)
  
  # Build formula
  formula_str <- paste("Spawners ~", paste(predictors, collapse = " + "))
  
  # Fit model
  model <- lm(as.formula(formula_str), data = river_data)
  
  # Backward elimination
  cat("Initial Model:\n")
  print(summary(model))
  
  # Calculate VIF if possible
  if (length(predictors) > 1) {
    cat("\n--- Variance Inflation Factors ---\n")
    tryCatch({
      vif_vals <- vif(model)
      print(vif_vals)
    }, error = function(e) {
      cat("VIF calculation not possible\n")
    })
  }
  
  # Diagnostic plots
  par(mfrow = c(2, 2))
  plot(model, main = river_name)
  par(mfrow = c(1, 1))
  
  return(model)
}

# Analyze Umatilla River
umatilla_data <- merged_data %>% 
  filter(Location == "Umatilla River")

umatilla_predictors <- c( "Yearly.Rainfall", "colsnakedams")
umatilla_model <- analyze_river(umatilla_data, "Umatilla River", umatilla_predictors)

```
```{r}
# Analyze Umatilla River
minam_data <- merged_data %>% 
  filter(Location == "Minam River")

minam_predictors <- c("FishConsumption", "DissolvedOxygen", "TempMax", "Yearly.Rainfall", "totaldams", "pH", "TempSummer", "Ammonia")
minam_model <- analyze_river(minam_data, "Minam River", minam_predictors)
```

# SECTION 8: LOGGING AND CONSUMPTION ANALYSIS
```{r}
cat("\n", rep("=", 80), "\n", sep = "")
cat("OVERALL SALMON POPULATION vs LOGGING AND CONSUMPTION\n")
cat(rep("=", 80), "\n\n", sep = "")

# Aggregate salmon by year
total_salmon <- merged_data %>%
  group_by(Year) %>%
  dplyr::summarise(
    TotalSpawners = sum(Spawners, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(logging_clean, by = "Year") %>%
  left_join(consumption_clean, by = "Year")

# Regression analysis
logging_model <- lm(TotalSpawners ~ TotalNWLogging + FishConsumption, 
                    data = total_salmon)
cat("Model: Total Spawners ~ Logging + Consumption\n")
print(summary(logging_model))

# Plot
plot(logging_model)
```

# SECTION 9: SUMMARY STATISTICS
```{r}

cat("Dataset dimensions:", nrow(merged_data), "rows x", ncol(merged_data), "columns\n\n")

cat("Years covered:", min(merged_data$Year, na.rm = TRUE), "-", 
    max(merged_data$Year, na.rm = TRUE), "\n\n")

cat("Number of unique locations:", n_distinct(merged_data$Location), "\n\n")

cat("Summary of spawner counts:\n")
print(summary(merged_data$Spawners))
```